---
title: "Description and usage of MsBackendWeizMass"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of MsBackendWeizMass}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,BiocStyle,RSQLite}
bibliography: references.bib
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendWeizMass")`<br />
**Authors**: `r packageDescription("MsBackendWeizMass")[["Author"]] `<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(BiocStyle)
```

# Introduction

The `Spectra` package provides a central infrastructure for the handling of Mass
Spectrometry (MS) data. The package supports interchangeable use of different
*backends* to import MS data from a variety of sources (such as mzML files). The
`MsBackendWeizMass` package allows import and handling MS/MS spectrum data from
WeizMass spectral library databases [@shahaf_weizmass_2016]. This enables
integration of the high quality WeizMass MS/MS spectral library into
`Spectra`-based annotation workflows [@rainer_modular_2022].


# Installation

The package can be installed with the `BiocManager` package. To install
`BiocManager` use `install.packages("BiocManager")` and, after that,
`BiocManager::install("RforMassSpectrometry/MsBackendWeizMass")` to install this
package.


# Accessing MS/MS data from a WeizMass database

To use the `MsBackendWeizMass` package, access to a WeizMass database (i.e. a
WeizMass MySQL database) is required. Connection information needs to be
requested from the original authors from the WeizMass database
[@shahaf_weizmass_2016]. In this section we use a tiny SQLite-based test
database which is included within this package that has the same database layout
then the WeizMass v2 database.

Below we load all required libraries and get the file name of the SQLite
database.

```{r load-libs}
library(Spectra)
library(MsBackendWeizMass)
library(RSQLite)
db <- system.file("sqlite", "weizmassv2.sqlite", package = "MsBackendWeizMass")
con <- dbConnect(SQLite(), db)
```

A `Spectra` object representing the data from the WeizMass database can be
created with the `Spectra` function providing the connection to the database as
well as specifying the backend to be used (`MsBackendWeizMass`).

```{r}
sps <- Spectra(con, source = MsBackendWeizMass())
sps
```

The `spectraVariables` function can be used to get all available *spectra
variables* from the database.

```{r}
spectraVariables(sps)
```

Individual spectra variables can be accessed using a dedicated function (such as
`rtime`, `msLevel`, etc), if available, or using the `$` operator. The
chemical formulas for the compounds of the spectra could for example be
retrieved using `$formula`:

```{r}
sps$formula
```

In addition it is possible to retrieve multiple spectra variables using the
`spectraData` function:

```{r}
spectraData(sps, c("rtime", "formula", "adduct"))
```

MS/MS peak data can be retrieved using the `peaksData` function which returns a
`matrix` with the values. Below we get thus the MS peaks for the first spectrum.

```{r}
peaksData(sps)[[1L]]
```

The WeizMass database provides also additional peak information that can be
listed using the `peaksVariables` function:

```{r}
peaksVariables(sps)
```

We could thus also query the annotations for the individual peaks by requesting
in addition the column `"peak_annotation"` in the `peaksData` call.

```{r}
peaksData(sps, c("mz", "intensity", "peak_annotation"))[[1L]]
```

Note however that, since `peaksData` **always** returns a `matrix`, also the m/z
and intensity values are reported as characters instead of numeric values. Thus,
it is advisable to query m/z and intensity values separately from peak
annotations.

Peak annotations can alternatively also directly extracted from a `Spectra`
object using the `$` operator.

```{r}
sps$peak_annotation
```

Note also that precursor m/z values are stored as character values in the
database, but converted to numeric by the backend during the data retrieval. For
stored values that can not be converted to a numeric an `NA` is thus reported.


# Session information

```{r}
sessionInfo()
```
